A comparison on R tools for nonlinear least squares modeling

### 2023-03-07 RefactoringNLS/ComparingNLStools.Rmd
minor change
more about refactoring

Indexed parameters -- those whose names have subscripts on some element that
is part of the coefficient list -- 
are a useful idea to simplify model specification in some cases. However, they
are also a focus of trouble.

@@

## Issue: code structure and documentation for maintenance

The `nls()` code is structured in a way that inhibits both maintenance and improvement.
In particular, the iterative setup is such that introduction of Marquardt stabilization
is not easily available. To obtain performance when the code was written, a large portion
was written in C with consequent calls and returns that
severely complicate the code. Over time, R has become much more efficient on modern computers, and
the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
could be reduced by moving code entirely to R.

While R and its packages are generally very well documented for usage, that documentation
rarely extends as far as it might for code maintenance. 
The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
where seemingly trivial differences in things like representation of numbers or
base index for arrays lead to traps for unwary programmers.

# Goals and progress of our effort to improve `nls()`

## Code rationalization and documentation

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the features in base R or what is packaged in `nlspkg` as described 
  in @PkgFromRbase22. The R version can either serve as the operational code or
  as a reference version to facilitate the maintenance of C, Fortran, or other 
  versions of the software.

- to streamline the overall `nls()` infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as new features we or others
  would like to add.

- to explain what we do, either in comments, examples, or separate maintainer documentation.
  Since we are complaining about the lack of explanatory material for the current
  code, we feel it is incumbent on us to provide such material for our own work, and if
  possible for the existing code.

## Rationalization of formula specifications

We have seen that
`nls()` uses a different formula specification from the default if the 
`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
add `algorithm="plinear"` to the standard call, making errors more likely.

Nonlinear models, apart from those using the "plinear" algorithm, need all the parameters specified.
That is, the model formula should be written down as it would be computed.
Thus a linear model (which is valid input for nonlinear estimation) should
be specified `y ~ a * x + b` rather than `y ~ x`. 
Moreover, we want to avoid the partially linear parameters appearing 
in the result object as `.lin1`, `.lin2`, etc., where the actual position 
of parameters in the model is not explicit. We are unclear why these parameters
begin with a dot, that is, that they have the "hidden" indicator.

A possible workaround to allow consistency would be to specify the 
partially linear parameters when the `plinear` option is specified. For example, 
we could use `algorithm="plinear(Asym)"` while keeping the general model formula 
from the default specification. This would allow 
for the output of different `algorithm` options to be consistent. However, we have not
yet tried to code such a change, and welcome discussion from those working with
other packages using model formulas.

<!-- It is also of interest to consider code that will detect partial linearity,  -->
<!-- rather than expecting the user to provide the special structure and settings. Work in this -->
<!-- area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution -->
<!-- we can find. -->
<!-- Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin  -->
<!-- is about a class of models that have almost nothing in common -->
<!-- with the present issue. The algorithm in @Birkisson2013 will need investigating further, -->
<!-- but we suspect that it is complicated enough that it will not easily be adapted to  -->
<!-- a program like `nls()`. Possibly a simpler algorithm may be found. -->

<!-- ``` -->
<!-- y ~ x + z + x*z -->
<!-- y ~ a + b*x +c*z + d * (x*z)  # <-- fully specified -->
<!-- y ~ a*Fn1(b, c, x, z)/Fn2() -->

<!-- ``` -->

## Rationalization of indexed models

Indexed models clearly have a place in some areas of research. 
<!-- We need to be able to process models, as the example above, where the -->
<!-- model formula is  -->
We need to be able to process models such as 
`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)`, where `Strip` is the index. Data
`Length` and `Conc` are available for all observations and parameter `th` applies
to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
could be an experimental run number, or some other categorization.

Given that `[ ]` provides the R method to index arrays, the presence of square
brackets in a model
formula signals indexing. The question is how to perform the correct calculations.

<!-- While tedious (??maybe dummies or fastdummies packages make it less so?),  -->
<!-- we could set up the model as a sum of similar models with dummy -->
<!-- variables to turn "on" or "off" the particular case, transforming the model -->
<!-- parameters to simple, non-indexed ones (as does `nls()`). -->

Our view is that inclusion of the indexing **within** `nls()` or similar tools introduces 
complexity that could be avoided by a suitable wrapper function. The modeling
formula must be expanded to encompass all the index cases, likely with indexed
parameters renamed e.g., `a[2]` becomes `a2` as at present with `nls()`. However,
a wrapper could process output to reconstitute the indexing of parameters.

## Streamlining code

We are pessimistic that the overall structure of `nls()` can be streamlined due
to the entanglement of so many features with the complicated mix of R, C and
Fortran. Indeed, despite digging into the code over some months, we do not feel
confident that we sufficiently understand it nor that we could maintain it. On the other
hand, we do believe equivalent functionality can be built, but with at least
some differences in how the features will be accessed.

@@

## A testing package: NLSCompare

One of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

@@

# Strategic choices in nonlinear model estimation

>>>>

Indexed parameters -- those whose names have subscripts on some element that
is part of the coefficient list -- 
are a useful idea to simplify model specification in some cases. However, they
are also a focus of trouble. Typically we have some common parameters and 
others that are tied to particular cases.

@@

@@

In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

@@

## Should nls() be refactored?

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the features in base R or what is packaged in `nlspkg` as described 
  in @PkgFromRbase22. Our intent in providing a purely R version of the code was
  either to replace `nls()` or to serve as a reference version aiding maintenance 
  of C, Fortran, or other components of the software.

- to streamline the overall `nls()` infrastructure. By this we mean a 
  refactoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as new features we or others
  would like to add.

- to explain what we do, either in comments, examples, or separate maintainer 
  documentation. While R and its packages are generally very well documented for
  usage, that documentation
  rarely extends as far as it might for code maintenance. The paucity of such
  documentation is exacerbated by the mixed R/C/Fortran code base,
  where seemingly trivial differences in things like representation of numbers or
  base index for arrays lead to traps for unwary programmers.

A serious obstacle to changing `nls()` is that there are many legacy applications
and examples. Furthermore, with `nls()` as part of the base R distribution, users
are drawn to using it even when it is less than ideal. A possible way forward
is to

- use a packaged version of `nls()` to provide legacy features, with notification
  to users that there are other tools available

- if maintainers are willing, to merge and rationalize the capabilities of 
  `minpack.lm`, `nlsr` and `gslnls`

- to encourage the (separate) development of capabilities such as indexed
  parameter models if these are in demand.

This direction does not, of course, resolve the inconsistencies in model 
specification in formulas for nonlinear models, which undoubtedly is an issue
for other modeling tools.

A possible workaround to allow consistency for the `plinear` algorithm 
would be to specify the linear parameters when the `plinear` option is specified. 
For example, we could use `algorithm="plinear(Asym)"` while keeping the general 
model formula from the default specification. This would allow 
for the output of different `algorithm` options to be consistent. However, we 
have not yet tried to code such a change, and welcome discussion from those working
with other packages using model formulas.

## Future of nonlinear model estimation in R
